BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2015-10-22 Paper/BOAST.tex
Rewrote introduction with Thierry.

Porting and tuning HPC applications to new platforms is tedious and costly in
terms of human resources. Nonetheless, it is a very important aspect of the
Mont-Blanc project. Indeed, for the project, more than ten applications were
selected to be ported and optimized for the prototype platform.

Unfortunately, portability efforts are often lost when migrating to a new
architecture. Worse, code may lose maintainability because several versions of
some functionalities coexist, usually with a lot of duplication.

Thus productivity of porting and tuning efforts is low as a huge fraction of
those developments are never used after the platform they were intended for is
decommissioned.

Genericity of HPC codes is often limited. One of the reason is that producing
generic code in Fortran 90/95 is difficult as the language does not really fit
for it. Sometimes, adding genericity degrades performance as optimization
opportunities that come from over-specification are lost.

Functionality of HPC codes is tied to the previous point. Without genericity,
adding new functionalities can be quite costly.

@@

@@

\section{Motivating Example: OpenCL Laplace}
\label{sec:laplace}

>>>>

Porting and tuning HPC applications to new platforms is tedious and costly in term of human resources.
Many developer teams are more concentrated on solving the physical problems faster and better with new algorithms
and simulating new physics. 
This is why optimization even if it is important is a task which is sometimes neglected or hidden for many developers.
In the same way, the domain specific language (DSL) has exactly this goal,  the complete separation of the high-level code dedicated to simulate new physics
and the low-level one which is more dependent to the HPC architecture. 
Nevertheless, decoupling the high and low levels don't avoid the need of optimization which becomes more and more unavoidable 
with the new architectures based on many cores and accelerators.

Unfortunately, portability efforts of the low level i.e. computing kernels are often lost when migrating to a new
architecture. Worse, code may lose maintainability because several versions of
some functionalities coexist, usually with a lot of duplication.
Thus productivity of porting and tuning efforts is low as a huge fraction of
those developments are never used after the platform they were intended for is
decommissioned.

The primary goal of the metaprogramming framework BOAST is to simplify the optimization of the computing kernels for HPC
applications. The main idea is to use a high level language to define an API of computing kernels i.e. subroutines and
their related arguments, but also describe the way to optimize it (unrolling, vectorization, parametrization, templating, tiling).
BOAST will test automatically the many possibilities of optimizations and choose the best one after validation from a
reference version.
Then BOAST can generate pieces of Fortran, C, OpenCL or CUDA codes, test if for a specific target architecture.

With the definition of an API, BOAST can hide the details of the implementation at high level but also is the way to
factorize different kernels depending only on parameters as datatypes, the length of filters or the boundary conditions.
The maintainability and genericity of HPC codes is often limited. One of the reason is that producing
generic code in Fortran 90/95 is difficult as the language does not really fit
for it. Sometimes, adding genericity or functionalities can be quite costly and degrades performance as optimization
opportunities that come from over-specification are lost.
BOAST can provide a solution to handle genericity and add new functionalities at low levels.
For instance, the optimization of short convolutions depends strongly
on the length of the filters. This is why HPC codes using finite differences handle only one order. BOAST allows
handling any order and even more generalizing to another numerical scheme.

In this paper, we illustrate the optimization problems related to scientific computing with the Laplace equation.
Then we present our solution, BOAST, its architecture and its functionalities. We apply BOAST to three use cases, the
Laplace equation, short multi-dimensional convolutions to perform differential operators and finally
the wavelet transform which are used in a quantum chemistry code, BigDFT.
Then BOAST is applied to a full application, SPECFEM3D dedicated to the seismic wave propagation simulator.
80 GPU computing kernels have been ported improving considerably the maintanability and providing a single code base for
OpenCL and CUDA kernel versions.
Finally we compare BOAST with other related works.

% Porting and tuning HPC applications to new platforms is tedious and costly in
% terms of human resources. Nonetheless, it is a very important aspect of the
% Mont-Blanc project. Indeed, for the project, more than ten applications were
% selected to be ported and optimized for the prototype platform.
% 
% Unfortunately, portability efforts are often lost when migrating to a new
% architecture. Worse, code may lose maintainability because several versions of
% some functionalities coexist, usually with a lot of duplication.
% 
% Thus productivity of porting and tuning efforts is low as a huge fraction of
% those developments are never used after the platform they were intended for is
% decommissioned.
% 
% Genericity of HPC codes is often limited. One of the reason is that producing
% generic code in Fortran 90/95 is difficult as the language does not really fit
% for it. Sometimes, adding genericity degrades performance as optimization
% opportunities that come from over-specification are lost.
% 
% Functionality of HPC codes is tied to the previous point. Without genericity,
% adding new functionalities can be quite costly.

@@

The motivation of the BOAST project comes from the need of the HPC communities which have full and large applications
difficult to port and maintain for emerging architectures based on accelerators, longer vector lengths, or low power
processors.

@@

\section{Basic Example: OpenCL Laplace}
\label{sec:laplace}
